---
title: "Week 1 - Part 1: Basic R"
author: "Dr. Donghyung Lee - Modified by Dr. Ashouri"
date: "Jan 29, 2025"
output: 
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Basic Calculator Operations

You can perform basic arithmetic operations directly.
```{r}
2+3*4   # Basic arithmetic
        # Exponentiation `^`
log(34) # Logarithm (natural log or ln)
        # Log10 log10()
exp(10) # Exponential exp()
        # Square Root sqrt()
        # Absolute Value
        # Round Down floor()
        # Round Up ceiling()

?log
```
You may use `#` (comment symbol) to comment out lines of your code. If you want to comment/uncomment a few lines together, use Ctrl+Shift+C (Windows) or Command+Shift+C (Mac).

## Variable Assignment
Variables in R can store a variety of data types including numbers, characters, and more. Variables are assigned using the `<-` operator (left-arrow operator). (shortcut: "Alt" + "-" (windows) and "Option"+"-" (mac))
```{r}
x <- 23 #Ctrl+enter (windows) Command+enter (mac)
x=25
 
xx <- x^2
xx

test <- "Math"
test

test.score <- 93.24
test.score

# x <- x + 5
```

### Naming Conventions
Variable names should start with a letter and can contain letters, numbers, period(.) and underscore(_) symbol.
```{r}
test2 <- 2
test2
# 2test <- 3
# test_2 <- 3

```

## Data Types in R

### Vectors

Vectors can be created using the c() function or seq() for sequences.
```{r}
# c() function
?c
y <- c(1,3,4,5,3,10,4,3,2) 

y[2] # Extract the 2nd element using index `[n]`
     # Extract the 2nd to 5th elements
     # Extract the 2nd, 4th, 6th elements
     # Extract all but the 2nd element
     # Extract all but the 3rd through 5th elements
     # Assign 5 to the 3rd element

# seq() function
?seq
yy <- seq(from=0,to=6,by=0.5)
yy

# If you don't know the function name google for "R compute function name (e.g., mean or standard deviation)"
mean(y)
sd(y)

# R uses the notation `[n]' at the beginning 
# of each output line to indicate which entries are shown on that line.
test.subjects <- c("Economics","Math","English")
test.subjects
```

### Vector Types
```{r}
# Numeric Vector (identical to double)
z <- 1:4
z
class(z)

zz <- c(1.3, 2.2, 3.3)
class(zz)

# Character Vector
w <- c("a","b","c")
class(w)

Grade <- c("freshman", "sophomore", "junior", "senior")
class(Grade)

# Logical Vector
h <- c(TRUE, TRUE, FALSE)
class(h)
TRUE == 1
FALSE == 0

treatment <- c("A","B","C","A","B","C")
class(treatment)

video <- c(47, 63, 58, 53, 53, 63, 53, 39, 58, 50)
video
video > 55 # Is the value of each element greater than 55?
vv <- video > 55
vv

# interpret the output from the following command
sum(video > 50)

# [Practice] 
# How can we get R to calculate the mean of only those elements
# of scores that are above 55?

```

### Matrices

Matrices in R must have elements of the same type.
```{r}
mat <- matrix(c(seq(1,17,1),rep(2,3)),nrow=4) # filled by columns
mat

mat2 <- matrix(c(seq(1,17,1),rep(2,3)),nrow=4,byrow=TRUE) # filled by rows
mat2

mat2[2,3] # Access element at row 2, column 3 using a [row,col] index pair
mat2[,2]
mat2[1,]
class(mat2[,4])

# However, matrix could not have a mix of numeric and characteristic columns. Only supports one type of object. 
mat3 <- matrix(c(seq(1,17,1),rep("2",3)),nrow=4) ## it turns all the objects to character
mat3
class(mat3[,2])
```

### Arrarys

Arrays are multi-dimensional data structures that are conceptually similar to matrices, but they can have more than two dimensions.
```{r}
array1 <- array(1:8,c(2,4))
array1

array("1",c(2,4))

array2 <- array(seq(1,30,1),c(2,3,5))
array2
array2[1,1,1]
```

### Data Frames

Data frames are similar to matrix but they allow for columns of different types.

```{r}
matrix(c(seq(1,3,1),c("A","B","C")),nrow=3)

test.data <- data.frame(Subjects=c("Economics","Math","English"),
                        Registered=c(TRUE,TRUE,FALSE),
                        Score=c(89,93,NA))
test.data
str(test.data) # display the structure of an arbitrary R object
class(test.data$Registered)
class(test.data$Score)
class(test.data$Subjects)
```


### Factor 

Factors are used for categorical variables and can be nominal (no order) or ordinal (ordered).

Categorical variable: 

1. without order --nominal -- Red, Blue, Orange                       
2. with order --  ordinal -- Freshman Sophomore, Junior,Senior
https://online.stat.psu.edu/stat200/lesson/1/1.1/1.1.1

                     
Refer to this website for more details

* Berkeley:https://www.stat.berkeley.edu/~s133/factors.html
* UCLA: https://stats.idre.ucla.edu/r/modules/factor-variables/

```{r}
# Nominal factor
color.vec <- c("Red","Blue","Orange","Blue","Red")
class(color.vec)
color.fac <- factor(color.vec)
color.fac
```

* Factors have levels, which define the set of possible values the factor can have.
* The levels in the factor variable are sorted alphabetically based on the unique elements found in 'color.vec'.

```{r}
factor(color.vec,levels=c("Red","Blue","Orange"))
# Change the factor labels
x <- c(1,2,4,2,4,2,5,3,2,3,4,1)
class(x)
factor(x)
factor(x,labels=c("A","B","C","D","E"))

# change both levels and labels
factor(x,labels=c("A","B","C","D","E"), levels = c(1, 3, 4, 2, 5))
```

```{r}
# Ordinal factor
ordered(color.vec)
ordered(color.vec,levels=c("Red","Blue","Orange"))
```

Advantages of using factor variables:

1. Useful in statistical models, DF can be computed correctly.
2. Useful in different types of graphics.
3. Factors are integers internally. Storing string variables as factor variables is more memory-efficient. 






